A Quantitative Genetic Linked Selection Model fit to the Negative Selection Process in Humans

### 2023-07-03 manuscript/manuscript.tex
minor ms changes
rework secs

In this work, we extend another class of linked selection models that derive
from quantitative genetics to address limitations of classic background
selection theory. These models suggest that the reduction in pairwise
coalescence times at a position in the genome is determined by the
recombination map and distribution of polygenic additive fitness variance along
the chromosome \parencite{Santiago1998-bs}. This reduction is due to cumulative
effect of stochastic perturbations from selection at other sites. While these
models can theoretically accommodate polygenic fitness variation from any
source as long as its rate of change is not too rapid, we focus specifically on
a deleterious-mutations-only model from Santiago and Caballero (2016). This
model is identical to the classic BGS model when selection against deleterious
mutations is strong, but it also correctly predicts the reduction in diversity
and substitutions rates of weakly deleterious mutations by accurately modeling
levels of genic fitness variation. We extend Santiago and Caballero’s
(hereafter the SC16) model so that it can be used to model genome-wide patterns
of diversity under negative selection processes, and develop Python software to
use this theoretic approach to fit genome-wide models of negative selection.
Using forward simulations, we show this model leads to more accurate DFE
estimates under weak selection. While this new model accommodates the
possibility that a given reduction B could be due to weaker selection, in
applying our composite-likelihood method to human population genomic data, we
confirm earlier findings that strong negative selection at a small fraction of
sites fits human data across populations best. Finally, we show that while this
model solves some limitations of the classic BGS model, simulations reveal
another mysterious defect: a poor fit around $Ns=1$. We show that this can be
partially corrected by incorporating the effects of selective interference on
genic variation, but the remaining discrepancy is due to the build up of
negative linkage disequilibrium that cannot be adequately captured by this
model. We find accounting for genic selective interference between segments
fits human population genomic data slightly better, suggesting levels of
negative selection in humans are sufficient to perturb the selection at other
sites.

@@

Overall, we find the PhastCons Priority and CADD 6\% models fit about equally
as well (Figure \ref{fig:figure-2}A), explaining $R^2=68.45$\% and
$R^2=68.04$\% of the out-sample variance in Yoruba pairwise diversity at the
megabase-scale, respectively. In populations impacted by the out-of-Africa
bottleneck, the goodness-of-fit was lower across all models (e.g. 62.98\% and
59.98\% for CEU and CHB respectively in the PhastCons Priority model). The CADD
6\% model was previously found by \textcite{Murphy2022-sj} to be the
best-fitting model using classic BGS. Our PhastCons Priority model fits
marginally better than the CADD 6\% model consistently across all three
populations (Supplementary Table \ref{supp:tbl-r2}). This is likely because the
PhastCons Priority model has more DFE parameters, and thus can fit finer-scale
differences in the DFE across its four feature types. By contrast, the CADD 6\%
track must fit patterns of diversity to conserved sites under a single DFE. The
CADD 8\% model fit slightly worse than the CADD 6\% across all populations, and
the Feature Priority model had the poorest-fit.

@@

Since our $R_\text{LOCO}^2$ estimates suggest our models fit left-out
chromosomes well even though our method assumes constant demography and
homogeneous mutation rates along the genome, we wondered what $R^2$ we could
expect under a simple theoretic model where $\cov_i(B_i', \varepsilon_i) = 0$
and $\var_i(\varepsilon_i)$ was determined by the variance in neutral
coalescence times alone with an effective population size rescaled by $B_i'$.
We calculate this ``coalescence $R^2$" by calculating the theoretic coalescence
time variance in a megabase window with recombination and an effective
population size of $B_i' N_e$ where we use our best-fitting model's as a
plug-in estimator of $B_i'$. We note that this is a rough approximation, since
this calculation assumes constant demography and that the variance in
coalescence times under negative selection is likely lower than expected under
drift alone.

Still, we find that our out-sample $R_\text{LOCO}^2$ for the Yoruba samples
($R_\text{LOCO}^2 \approx 68$\%) is close to the theoretic $R_\text{coal}^2
\approx 67$\%. For bottlenecked out-of-Africa populations, $R_\text{coal}^2
\approx 64$\% compared to the observed $R_\text{LOCO}^2 \approx 62$\% for CEU
and $R_\text{LOCO}^2 \approx 59$\% for CHB. We note that because bottlenecks
increase the variance in coalescence times beyond the level implied by the
effective population size, the disagreement between observed $R_\text{LOCO}^2$
and the theoretic $R_\text{coal}^2$ is likely to shrink. Overall, this suggests
that negative selection models explain the vast majority coalescence time
variation at the megabase-scale that is capable of being explained (i.e. that
is not coalescence noise).

@@

Linked selection theory suggests the reduction signal is proportional to the
spatial distributions of recombination and fitness variance along the
chromosome.

Next, we developed a composite-likelihood method to estimate expected diversity
without linked selection $\pi_0$, the deleterious mutation rate ($\mu_D$), and
distribution of fitness effects ($\mathbf{W}$) under negative selection.
Because the focus of our work is to fix deficiencies in the classic BGS model,
our current implementation does not consider reductions in diversity due to
recurrent hitchhiking or other positive selection processes. Since previous
work has established that purifying selection is the dominant mode of linked
selection in humans and hitchhiking plays a limited role
\parencite{McVicker2009-ax,Murphy2022-sj}, humans are a suitable organism to
test our method.

>>>>

In this work, we extend another class of linked selection models that derive
from quantitative genetics to address limitations of classic background
selection theory. These models suggest that the reduction in pairwise
coalescence times at a position in the genome is determined by the
recombination map and distribution of polygenic additive fitness variance along
the chromosome \parencite{Santiago1998-bs}. This reduction is due to cumulative
effect of stochastic perturbations from selection at other sites. While these
models can theoretically accommodate polygenic fitness variation from any
source as long as its rate of change is not too rapid, we focus specifically on
a deleterious-mutations-only model from Santiago and Caballero (2016). This
model is identical to the classic BGS model when selection against deleterious
mutations is strong, but it also correctly predicts the reduction in diversity
and substitutions rates of weakly deleterious mutations by accurately modeling
levels of genic fitness variation. We extend Santiago and Caballero’s
(hereafter the SC16) model so that it can be used to model genome-wide patterns
of diversity under negative selection processes, and develop Python software to
use this theoretic approach to fit genome-wide models of negative selection.
Using forward simulations, we show this model leads to more accurate DFE
estimates under weak selection. While this new model accommodates the
possibility that a given reduction B could be due to weaker selection, in
applying our composite likelihood method to human population genomic data, we
confirm earlier findings that strong negative selection at a small fraction of
sites fits human data across populations best. Finally, we show that while this
model solves some limitations of the classic BGS model, simulations reveal
another mysterious defect: a poor fit around $Ns=1$. We show that this can be
partially corrected by incorporating the effects of selective interference on
genic variation, but the remaining discrepancy is due to the build up of
negative linkage disequilibrium that cannot be adequately captured by this
model. We find accounting for genic selective interference between segments
fits human population genomic data slightly better, suggesting levels of
negative selection in humans are sufficient to perturb the selection at other
sites.

@@

Overall, we find the PhastCons Priority and CADD 6\% models fit about equally
as well (Figure \ref{fig:figure-2}A), explaining $R^2=68.45$\% and
$R^2=68.04$\% of the out-sample variance in Yoruba pairwise diversity at the
megabase-scale, respectively. Generally, our models predict out-sample
diversity levels slightly better than previous methods; the best-fitting CADD
6\% model from \textcite{Murphy2022-sj} explained 60\% of diversity in left-out
2 Mbp windows across YRI samples.

In populations impacted by the out-of-Africa bottleneck, the goodness-of-fit
was lower across all models (e.g. 62.98\% and 59.98\% for CEU and CHB
respectively in the PhastCons Priority model). Our PhastCons Priority model
fits marginally better than the CADD 6\% model consistently across all three
populations (Supplementary Table \ref{supp:tbl-r2}). This is likely because the
PhastCons Priority model has more DFE parameters, and thus can fit finer-scale
differences with the four feature types. By contrast, the CADD 6\% track must
fit patterns of diversity to conserved sites under a single DFE. The CADD 8\%
model fit slightly worse than the CADD 6\% across all populations, and the
Feature Priority model had the poorest-fit.

@@

Overall, our $R_\text{LOCO}^2$ estimates suggest our negative selection model
explains up to 68\% of out-sample variance in diversity of the megabase scale,
even though our method assumes constant demography and homogeneous mutation
rates along the genome. A worthwhile question is: how much variation
\emph{could} we expect to fit at this scale? Given that selection alters
genealogies in ways beyond just decreasing mean pairwise coalescence time and
populations have non-constant demography, an exact analytic answer is
intractable. However, we can get an approximate idea if we assume that the
residual variance $\sum_b (\pi_b - \widehat{\pi}_b)^2$ is determined entirely
by the expected neutral coalescence noise around the expected coalescence time
$2B(b)N$ (see Method XXX). Thus, we define $R_\text{coal}^2$ as the theoretic
proportion of variance explainable under this model to ballpark the variance
capable of being explained.

We find that our out-sample $R_\text{LOCO}^2$ for the Yoruba samples
($R_\text{LOCO}^2 \approx 68$\%) is slightly above the theoretic
$R_\text{coal}^2 \approx 67$\%. This suggests our model is in the vicinity of
fitting all possible signal possible under these simplifying assumption. It
should be noted that selection is expected to \emph{decrease} the variance in
coalescence times beyond simply rescaling effective population size implies;
better future models for $R_\text{coal}^2$ could show there is more linked
selection signal to explain.

For bottlenecked out-of-Africa populations, we find a larger discrepancy
between $R_\text{coal}^2$ and observed out-sample $R_\text{LOCO}^2$. The
theoretic $R_\text{coal}^2 \approx 64$\% for both populations, compared to the
observed $R_\text{LOCO}^2 \approx 62$\% for CEU and $R_\text{LOCO}^2 \approx
59$\% for CHB. Given that bottlenecks would act to increase the residual
variance in coalescence times beyond the level implied by the effective
population size, the gap would likely shrink under more realistic models or
simulation-based approximations for $R_\text{coal}^2$. Overall, this suggests
that negative selection models explain the vast majority coalescence time
variation at the megabase-scale that is capable of being explained (i.e. that
is not coalescence noise).

@@

Our composite likelihood method has three sets of parameters: expected
diversity in the absence of linked selection diversity $\pi_0$, the mutation
rate $\mu$, and the distribution of fitness effects $\mathbf{W}$ across the
selection grid for each of the $K$ feature types. Given that the relationship
between $\pi$ and the strength selection is U-shaped, we wondered whether the
observed genome-wide patterns of diversity could fit under weaker selection
than found in previous methods based on classic BGS theory.

However, across all of our annotation models, new deleterious mutations in
putatively conserved features are consistently estimated to have strongly
deleterious effects (Figure \ref{fig:figure-3}A). This includes features such
as coding sequences, PhastCons regions, or the top 6\% of basepairs with the
most pathogenic CADD scores. Thus, even with a model that can better
distinguish weak versus strong selection, our DFE estimates suggest

qualitatively
agree with recent work \parencite{Murphy2022-sj}.

Moreover, the estimated mutation rates are close
to pedigree-based estimates for most of these annotation models (XXX).

Second, we find evidence that our method will indeed estimate weak selection
across features

Given that the B' method more
accurately predicts diversity across weaker selection coefficients, we wondered
whether our inferred DFEs could

Linked selection theory suggests the reduction signal is determined by the
spatial distributions of recombination and fitness variance along the
chromosome. When fitness variation is determined mostly by negative selection,
the

Where does the information for the DFE come from?

All of our models show a general trade off in fitting the linked selection
signal.

Next, we developed a composite likelihood method to estimate expected diversity
without linked selection $\pi_0$, the deleterious mutation rate ($\mu_D$), and
distribution of fitness effects ($\mathbf{W}$) under negative selection.
Because the focus of our work is to fix deficiencies in the classic BGS model,
our current implementation does not consider reductions in diversity due to
recurrent hitchhiking or other positive selection processes. Since previous
work has established that purifying selection is the dominant mode of linked
selection in humans and hitchhiking plays a limited role
\parencite{McVicker2009-ax,Murphy2022-sj}, humans are a suitable organism to
test our method.
